Northern Pikeminnow Abundance in Deadwater Slough and Potential Impacts to Local Chinook Salmon Populations

### 2021-11-04 analysis/paper/paper.Rmd
additions to Introduction

---
title: "Northern Pikeminnow Abundance in Deadwater Slough and Potential Impacts to Local Chinook Salmon Populations"
author:
  - Richard A. Carmichael:
      email: Richard.Carmichael@merck.com
      institute: [biomark]
      correspondence: true
  - Nicholas J. Porter:
      email: Nicholas.Porter@merck.com
      institute: [biomark]
      correspondence: false
  - Kevin E. See:
      email: Kevin.See@merck.com
      institute: [biomark]
      correspondence: false
  - Michael S. Hall:
      email: Mike.Hall1@merck.com
      institute: [biomark]
      correspondence: false
  - Michael W. Ackerman:
      email: Mike.Ackerman@merck.com
      institute: [biomark]
      correspondence: false
institute:
  - biomark: Biomark, Inc., 705 South 8th Street, Boise, Idaho, 83702, USA
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      number_sections: FALSE
      toc: no
      always_allow_html: true
      fig_width: 6
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
fontsize: 12pt
mainfont: Times New Roman
bibliography: 
- references.bib      # Kevin's library
- AckermanLibrary.bib # Mike's library
csl: "../templates/american-fisheries-society.csl" # Insert path for the bib-style
abstract: |
  Abstract text here. 
keywords: |
  northern pikeminnow; Chinook salmon; predation; mark-recapture; bioenergetics
highlights: |
  These are the highlights. 
---

@@

***Paragraph 1***

* The Lemhi River supports an ESA-listed Chinook salmon population with substantial cultural and recreational value and is a candidate for protection and/or recovery.
* The population has become depleted with lower adult returns in recent years.
* Decreased returns and/or increased mortality of the population have often been attributed to (in chronological order by life stage):
  - Degraded natal tributary habitat (e.g., high amount of agriculture)
  - Reduced survival in the mainstem corridor e.g., Snake and Columbia rivers
  - Poor ocean conditions
  - Others worth mentioning? Perhaps adult harvest in ocean and/or in-river returns?

@@

* One potentially under-appreciated source of mortality on juvenile Chinook salmon is predation from piscivorous fishes, native and/or non-native
  - At this point, I'm thinking leave avian predation out of the equation, but instead include in Discussion
  - and keep focus on the freshwater environment
* What is known about downriver (Columbia/Snake) piscivorous fish populations and potential predation on juvenile Chinook salmon?
  - literature review of northern pikeminnow, smallmouth bass, other? predation on juvenile Chinook salmon, particulary from the Snake and Columbia rivers
  - still stay away from avian predation

@@

* Chinook salmon in the Lemhi River exhibit two juvenile migration tactics (@Copeland2014c and See et al. *in prep*).
  - downstream rearing (DSR)
  - natal reach rearing (NRR)
* Define these juvenile migration tactics for stream-type Chinook salmon

>>>>

---
title: "Northern Pikeminnow Abundance in Deadwater Slough and Potential Impacts to Local Chinook Salmon Populations"
author:
  - Richard A. Carmichael:
      email: Richard.Carmichael@merck.com
      institute: [biomark]
      correspondence: true
  - Nicholas J. Porter:
      email: Nicholas.Porter@merck.com
      institute: [biomark]
      correspondence: false
  - Kevin E. See:
      email: Kevin.See@merck.com
      institute: [biomark, wdfw]
      correspondence: false
  - Michael S. Hall:
      email: Mike.Hall1@merck.com
      institute: [biomark]
      correspondence: false
  - Michael W. Ackerman:
      email: Mike.Ackerman@merck.com
      institute: [biomark, mhe]
      correspondence: false
institute:
  - biomark: Biomark, Inc., 705 South 8th Street, Boise, Idaho, 83702, USA
  - wdfw: Washington Department of Fish and Wildife, *Address Here*
  - mhe: Mount Hood Environmental, PO Box 4282, McCall, Idaho, 83638, USA
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      number_sections: FALSE
      toc: no
      always_allow_html: true
      fig_width: 6
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
fontsize: 12pt
mainfont: Times New Roman
bibliography: 
- references.bib      # Kevin's library
- AckermanLibrary.bib # Mike's library
csl: "../templates/american-fisheries-society.csl" # Insert path for the bib-style
abstract: |
  Abstract text here. 
keywords: |
  northern pikeminnow; Chinook salmon; predation; mark-recapture; bioenergetics
highlights: |
  These are the highlights. 
---

@@

The Upper Salmon River major population group (MPG) supports eight extant spring/summer Chinook salmon *Oncorhynchus tshawytscha* populations including Salmon River (above Redfish Lake), Valley Creek, Yankee Fork Salmon River, East Fork Salmon River, Salmon River (mainstem below Redfish), Pahsimeroi River, Lemhi River, and North Fork Salmon River [@NOAA2017]. For the MPG to be considered viable, at least five of these eight populations should meet viability criteria set forth by @McElhany2000 and @ICTRT2007, and is necessary for recovery of the Snake River Evolutionary Significant Unit (ESU). Populations within the ESU have substantial cultural value, support downriver mainstem Snake and Columbia River commercial and subsistence fisheries, and in year with sufficient abundance, support local fisheries that contribute to the local economies. All Upper Salmon River MPG populations have become depleted in recent decades. Declines in survival of juvenile Chinook salmon have been attributed to the removal of beavers from the landscape (fur trade), mining activities, simplification of rivers resulting from the construction of railroads and agriculture, water withdrawals, logging activities, urbanization, avian predation, proliferation of non-native species (e.g., non-native coastal rainbow trout *O. mykiss irideus* and brook trout *Salvelinus fontinalis*), warming streams and rivers, and modifications to downriver migration corridors (e.g., from hydropower projects). Moreover, the abundance of adult returns are further impacted by ocean and downriver harvests, altered ocean conditions, and changes to the spawning migration corridor. Each of these factors have contributed, to varying and unknown extents, to reduced adult escapement, the primary metric used to assess population viability. In response to the above human activities and reduced abundances, action agencies have invested substantial funds in recent years to rehabilitate tributary ecosystems in an attempt to improve juvenile survival and adult spawning conditions.

@@

One potentially important, but perhaps under-appreciated source of mortality on Chinook salmon is predation on emigrating juveniles by piscivorious fishes, including both native and non-native species. As an example,...

  - At this point, I'm thinking leave avian predation out of the equation, but instead include in Discussion
  - and keep focus on the freshwater environment
* What is known about downriver (Columbia/Snake) piscivorous fish populations and potential predation on juvenile Chinook salmon?
  - literature review of northern pikeminnow, smallmouth bass, other? predation on juvenile Chinook salmon, particulary from the Snake and Columbia rivers
  - still stay away from avian predation

@@

Spring/summer Chinook salmon in the Upper Salmon MPG are stream-type and exhibit two distinct migration tactics, termed downstream rearing (DSR) and natal reach rearing (NRR) by [@Copeland2014]. The DSR migrants leave the natal spawning area as subyearlings during June through November and typically winter in downstream, mainstem habitats prior to emigration to the ocean the following spring as smolts. Alternatively, NRR migrants remain in their natal areas for approximately one year after emergence prior to emigration to the ocean as smolts. The diversity of emigration tactics provides for increases population resiliency due to changes in the freshwater rearing and migration environments and buffers against catastrophic events.
